Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market

### 2023-07-13 paper/final_submission/paper.tex
made final submission version splitting main text TEX file and supp PDF

To determine the relationship between the 696 FASTQ files released by the Chinese CDC on the NGDC on March-29-2023 and the 227 files downloaded earlier that month by \citet{crits2023genetic}, I computed the SHA-512 hashes for all the NGDC FASTQ files and compared them to the hashes reported in Table S1 of \citet{crits2023genetic}.
This analysis confirmed that the dataset released on the NGDC contained unmodified versions of all 227 FASTQ files downloaded by \citet{crits2023genetic}, plus an additional 469 FASTQ files (Tables~\ref{tab:sha526} and \ref{tab:metadata}).

\subsection{Analysis of mitochondrial sequences validates mammalian metagenomics by Crits-Christoph et al, and also finds abundant non-mammalian chordates}
To determine the metagenomic content of the samples, I performed an analysis that conceptually parallels that described by \citet{crits2023genetic}.
Throughout this paper, I report results only for sequencing described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the small number of samples that involve amplicon-based sequencing of SARS-CoV-2 or RNA sequencing for viral whole-genome assembly (Table~\ref{tab:metadata}).

Briefly, the deep sequencing data were pre-processed to retain only high-quality reads.
These reads were aligned to the SARS-CoV-2 genome and a representative set of chordate mitochondrial genomes.
%The set of mitochondrial genomes was assembled by downloading all NCBI RefSeq genomes, filtering for those from chordates, and removing genomes that were highly similar to other retained genomes (within a Mash distance of $<$0.07).
This set included all the mammalian mitochondrial genomes from all species for which alignments were reported in \citet{crits2023genetic}.
The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table~\ref{tab:mito_retained}.
The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.

I compared the mitochondrial compositions of the samples determined by my analysis to those reported by \citet{crits2023genetic}.
Note that while \citet{crits2023genetic} describe aligning the sequencing data to all metazoa mitochondria, they only report alignment counts for mammals.
The aligned read counts for mammals from my analysis are highly correlated with those reported by \citet{crits2023genetic} for all sequencing runs with a reasonable number of total counts (Figure~\ref{fig:crits_christoph_vs_current_run_corr}).
My analysis and that of \citet{crits2023genetic} are independently implemented, and use different alignment programs, reference genome sets, and quality filtering criteria---so the fact that the two analyses give highly correlated results is reassuring.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figures/composition.pdf}
\caption{
Metagenomic composition of sample Q61.
{\bf (A)}
Composition as determined by aligning reads to mitochondrial genomes.
From left to right: composition determined in current study across all chordates, composition determined in current study across all mammals, composition reported in \citet{crits2023genetic} for just mammals shown in the first figure of their report, and composition reported \citet{crits2023genetic} across all mammals in the third supplementary table of their report.
{\bf (B)}
Composition determined in the current study by aligning assembled contigs to the four indicated genomes; this composition is similar to that reported in the third figure of \citet{crits2023genetic}.
See \url{https://jbloom.github.io/Huanan_market_samples/mito_composition.html} for an interactive version of (A) that allows similar pie charts to be viewed for any sample.
See \url{https://jbloom.github.io/Huanan_market_samples/genomic_contig_composition.html} for an interactive version of (B).
\label{fig:composition}
}
\end{figure*}

\subsection{Most environmental samples contain little or no SARS-CoV-2 reads}
To quantify the SARS-CoV-2 content of the samples, I plotted the percentage of all high-quality reads that aligned to SARS-CoV-2 (Figure~\ref{fig:sars2_aligned}, Table~\ref{tab:sars2_aligned_by_run}, Table~\ref{tab:sars2_aligned_by_sample}).
Over three-quarters of the samples had no reads aligning to SARS-CoV-2, and most of the rest only had a small number of viral reads.

@@

\begin{figure*}
\centerline{\includegraphics[width=0.85\linewidth]{figures/sars2_aligned.pdf}}
\caption{
Percentage of high-quality reads that align to SARS-CoV-2 for each sample.
Points are colored according to the date that the sample was collected.
Note the x-axis uses a symlog scale.
See \url{https://jbloom.github.io/Huanan_market_samples/sars2_aligned_vertical.html} for an interactive version of this plot where you can mouseover points for details including the mitochondrial composition of each sample, and select only samples from specific dates or from locations.
\label{fig:sars2_aligned}
}
\end{figure*}

@@

See Table~\ref{tab:susceptible_mammal} for a similar table that shows mammalian rather than chordate composition.
This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table~\ref{tab:raccoon_dog_long} (for raccoon dog) and Table~\ref{tab:mito_composition_by_sample} (for all species).

@@

 Overall, there were SARS-CoV-2 reads in just 3 of the 28 samples with at least 20\% of their chordate mitochondrial composition from the non-human susceptible species thought to have been sold live in the market (Table~\ref{tab:susceptible}).
There were 14 samples with at least 20\% chordate mitochondrial composition from raccoon dogs, but only sample Q61 contained any SARS-CoV-2 reads (1 of $\sim2.1 \times 10^8$ high-quality reads).
None of the six samples with at least 20\% chordate mitochondrial composition from bamboo rats contained any SARS-CoV-2 reads.
There was one sample each with at least 20\% chordate mitochondrial composition from Amur hedgehog and Malayan porcupine that contained SARS-CoV-2 reads (Table~\ref{tab:susceptible}).
If mitochondrial composition is instead analyzed only among mammals rather than chordates, Q61 is still the only sample with at least 20\% raccoon dog composition that contains any SARS-CoV-2 reads (Table~\ref{tab:susceptible_mammal}).
There is one sample with at least 20\% of its mammalian (but not chordate) mitochondrial composition from bamboo rat that contains a small number of SARS-CoV-2 reads (Table~\ref{tab:susceptible_mammal}).

Note that the 20\% cutoff applied in Tables~\ref{tab:susceptible} and \ref{tab:susceptible_mammal} is not integral to the analysis, but is just a way to subset on samples containing the most material from the species of interest to make the tables small enough to easily visualize.
See Tables~\ref{tab:raccoon_dog_long} and \ref{tab:mito_composition_by_sample} for much larger tables that show comparable data for all samples with respect to the raccoon dog or overall mitochondrial composition, respectively.
For instance, Table~\ref{tab:raccoon_dog_long} shows that there are a few samples with lower raccoon dog mitochondrial composition that also contain some SARS-CoV-2 reads.
However, ultimately these very large tables are difficult to visualize, and so the relationship between SARS-CoV-2 and mitochondrial content for each species for all samples is probably more facilely visualized using the interactive versions of the scatter plots described in the next subsection.

@@

Furthermore, the correlations are highly contingent on the sample set and details of how the statistics are calculated.
For instance, if we exclude the first sampling timepoint and only look at samples calculated on January-12-2020 or later, then the most correlated species is the Oriental rat snake, with the correlation for humans becoming even weaker and that for raccoon dogs remaining negative (Figure~\ref{fig:per_species_corr_faceted_later_dates}).
The correlations also change if only consider samples containing a non-zero number of SARS-CoV-2 reads, or calculate correlations on a linear rather than log scale (Figure~\ref{fig:different_corrs}).
The interactive versions of the correlation plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} and \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} enable the reader to explore the correlations for different sample subsets and methods of calculating the statistics.
%Exploration of these interactive plots shows that it takes a great deal of cherry picking of parameter combinations to put any plausibly infectable hosts at the top of the list of species whose mitochondrial material is most correlated with the abundance of SARS-CoV-2.

\begin{figure*}
\includegraphics[width=\linewidth]{figures/per_species_corr_faceted.png}
\caption{
Correlation between percent of all reads mapping to SARS-CoV-2 and the mitochondrial genome of each of the indicated species.
Each point represents a different environmental sample, and the orange text shows the Pearson correlation.
The scales are log10, and values of zero (which cannot be plotted on a log scale) are shown as half the minimum non-zero value observed across all samples.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot that enables mouseover of points for sample details, selection only of samples collected on specific dates or containing at least one SARS-CoV-2 reads, adjustment of scales from log to linear, and adjustment of mitochondrial percent to be of reads mapping to any mitochondria rather than of all reads.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for similar plots for individual species.
The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
\label{fig:per_species_corr_faceted}
}
\end{figure*}

@@

\begin{figure}
\includegraphics[width=\linewidth]{figures/overall_corr/overall_corr.pdf}
\caption{
Correlations between SARS-CoV-2 content and mitochondrial content for all species.
Top row show just samples containing at least one SARS-CoV-2 read and bottom row shows samples regardless of SARS-CoV-2 content; left shows samples from all sampling dates and right shows just samples from the January-12-2020 date when most of the wildlife sampling occurred.
This plot is designed to mimic the fourth figure of \citet{liu2022surveillance}.
Some key species are labeled; see the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
See Figure~\ref{fig:theil_sen_corr} for a version of this plot that uses correlations based on a Theil-Sen estimator that is more robust to outliers.
\label{fig:overall_corr}
}
\end{figure}

To remedy this omission, I used the chordate mitochondrial compositions calculated in the current study to generate plots analogous to that in the original Chinese CDC preprint.
When the correlations are taken across all sampling dates only for samples containing at least one SARS-CoV-2 read (as done in the original Chinese CDC preprint), then the genetic material of several non-human species (largemouth bass, catfish, cow, sheep, pig) is more correlated with SARS-CoV-2 reads than is human material (Figure~\ref{fig:overall_corr}, upper left).
If the plot is expanded to all samples (regardless of whether or not they contain SARS-CoV-2 reads), the trends are broadly similar, with fish remaining the most correlated species (Figure~\ref{fig:overall_corr}, lower left).
It is also informative to take the correlations only across samples collected on January-12-2020, which was the date of most intensive sampling of the wildlife stalls (this is when Q61 was collected).
 For January-12-2020 samples that contain SARS-CoV-2 reads, humans are among the species whose genetic material is most correlated with SARS-CoV-2 reads, but goat and spotted dove have roughly equivalent correlations (Figure~\ref{fig:overall_corr}, upper right).
 If we include January-12-2020 samples regardless of SARS-CoV-2 content, then snakehead fish and Malayan porcupine are among the most correlated species.
Raccoon dog and bamboo rat genetic material are not positively correlated with SARS-CoV-2 reads in any of the sample sets.
See the interactive version of Figure~\ref{fig:overall_corr} at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to explore other sample subsets.
The trends are broadly similar if the correlations are instead calculated based on a Theil-Sen estimator~\citep{theilsen} that is more robust to outliers (Figure~\ref{fig:theil_sen_corr}).

@@

Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2 (Table~\ref{tab:rt-qpcr} gives sequencing and RT-qPCR data for all samples that were either sequenced or called positive by the Chinese CDC).
However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
This inconsistency is illustrated in Table~\ref{tab:consistency} for four example samples: F100 is clearly positive (it both tested positive by RT-qPCR and contained thousands of reads mapping to SARS-CoV-2), but there is no statistical rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative.

@@

For the current study, all mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
The Amur hedgehog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
All the mitochondrial genomes were then filtered to remove highly similar ones.
To do this, \texttt{Mash}~\citep{ondov2016mash} was used to compute the mash distances between all pairs of mitochondrial genomes.
To ensure the most relevant mitochondrial genomes were retained, I first specified for manual retention the genomes for all the species listed in the third supplementary file of \citet{crits2023genetic}, as well as genomes for some additional relevant species (these species are listed under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}).
After retaining all of these manually specified mitochondrial genomes, the pipeline then greedily iterated through all other genomes choosing for retention each genome that was not within a mash distance of 0.07 of another already retained genome.
The full set of retained mitochondrial genomes is listed in Table~\ref{tab:mito_retained}.

@@

I then used \texttt{CoverM} (\url{https://github.com/wwood/CoverM}) to compute the number of aligned reads for each run, requiring alignment lengths of at least 40 nucleotides with at least 95\% identity, and excluding the 100 nucleotides at the contig ends.
The resulting statistics on alignment counts to SARS-CoV-2 and the mitochondrial genomes for each sequencing run are in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:sars2_aligned_by_run}.
To get per-sample alignment counts, I aggregated counts for all runs for each sample to get the statistics in Tables~\ref{tab:mito_composition_by_sample} and \ref{tab:sars2_aligned_by_sample}.
For the mitochondrial genome compositions and all other analyses in this paper, I only included only the samples from metagenomic sequencing, which are annotated with the description ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' in the metadata provided by \citet{liu2023surveillance}.
Species were only retained in the annotated reference set if they had at least 20\% of all aligned mitochondrial reads and at least 4000 covered based for at least one run, or if they are one of the species specified under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}.
The species specified under that key include all susceptible species studied in \citet{crits2023genetic}, so this criteria does not lead too dropping of any of the species in their study.

@@

\newgeometry{left=1.2in, right=1.2in, top=48pt, bottom=60pt}
\onecolumn
\renewcommand{\thepage}{S\arabic{page}}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\clearpage

\begin{figure*}
\includegraphics[width=\linewidth]{figures/crits_christoph_vs_current_run_corr.pdf}
\caption{
Correlation of the number of read counts mapping to mitochondrial genomes of mammalian species between the current study and \citet{crits2023genetic} for each sequencing run shared across the studies.
The correlation is calculated only across species for which \citet{crits2023genetic} provide read counts in the third supplementary figure of their report, and only for runs for which both studies find $\ge$50 reads mapping to that set of species.
See \url{https://jbloom.github.io/Huanan_market_samples/crits_christoph_vs_current_run_corr.html} for an interactive version of this plot that enables mouseover of points for details on specific runs, and adjustment of the threshold for how many total aligned mitochondrial reads a run must contain to be shown in this plot.
\label{fig:crits_christoph_vs_current_run_corr}
}
\end{figure*}

\clearpage

\begin{figure*}
\includegraphics[width=\linewidth]{figures/per_species_corr_faceted_later_dates.png}
\caption{
Correlation between percent of all reads mapping to SARS-CoV-2 and the mitochondrial genome of each of the indicated species only among samples collected after the first (Jan-1-2020) sampling date.
Except for the fact that this figure excludes samples from the first date, this figure is otherwise the same as Figure~\ref{fig:per_species_corr_faceted}.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot with numerous options, including allowing further refinement of the sample date range.
The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
\label{fig:per_species_corr_faceted_later_dates}
}
\end{figure*}

\clearpage

\begin{figure*}
\includegraphics[width=\linewidth]{figures/different_corrs/correlations.pdf}
\caption{
The correlations between the amount of SARS-CoV-2 and mitochondrial genetic material in the samples depends on how the correlation is calculated.
The top row shows correlations for human, and the bottom row shows correlations for raccoon dog.
Each column shows the correlations calculated different ways: with a log or linear scale, calculating the mitochondrial percent to be of all reads or just reads mapping to the mitochondria, and showing all samples or only those with at least one SARS-CoV-2 read.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for interactive versions of the plot panels.
The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
\label{fig:different_corrs}
}
\end{figure*}

\clearpage

\begin{figure}
\centering
\includegraphics[width=0.6\linewidth]{figures/theil_sen_corr/theil_sen_corr.png}
\caption{
A version of Figure~\ref{fig:overall_corr} where the y-axis shows a correlation computed from the Theil-Sen estimator~\citep{theilsen} rather than the Pearson correlation.
The Theil-Sen estimator is more robust to outliers, so species where the correlation is driven by just a few species will have values of zero in these plots.
See \url{https://jbloom.github.io/Huanan_market_samples/theil_sen_corr.html} for an interactive version of this plot.
\label{fig:theil_sen_corr}
}
\end{figure}

\clearpage

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/crits_christoph_data/check_sha512_vs_crits_christoph.csv} for the correspondence between the FASTQ files analyzed by \citet{crits2023genetic} and the files deposited in the NGDC by \citet{liu2023surveillance}.
The files are matched based on their SHA-526 hashes, which were computed directly for the NGDC files as part of the current study, and were taken from Table S1 of \citet{crits2023genetic} for that study.
\label{tab:sha526}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/metadata/merged_metadata.csv} for the metadata for all of the samples and files uploaded to the NGDC by \citet{liu2023surveillance} under accession CRA010170.
\label{tab:metadata}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/mitochondrial_genomes/retained.csv} for the NCBI accessions of the set of chordate mitochondrial genomes that was used as the alignment reference for the metagenomic analysis.
\label{tab:mito_retained}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/mito_composition_by_run.csv} for the counts of reads aligning to each mitochondrial genome for each sequencing run.
\label{tab:mito_composition_by_run}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/mito_composition_by_sample.csv} for the counts of reads aligning to each mitochondrial genome for each sample.
\label{tab:mito_composition_by_sample}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/sars2_aligned_by_run.csv} for the counts of reads aligning to SARS-CoV-2 for each sequencing run.
\label{tab:sars2_aligned_by_run}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/sars2_aligned_by_sample.csv} for the counts of reads aligning to SARS-CoV-2 for each sequencing sample.
\label{tab:sars2_aligned_by_sample}
}
\end{table}
Reads mapping to SARS-CoV-2 out of all high-quality (pre-processed) reads for samples with $\ge$20\% of their mammalian mitochondrial composition from a susceptible non-human species as defined in \citet{crits2023genetic}.
See Table~\ref{tab:susceptible} for a similar table that shows chordate rather than mammalian composition.
This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table~\ref{tab:raccoon_dog_long} (for raccoon dog) and Table~\ref{tab:mito_composition_by_sample} (for all species).
\label{tab:susceptible_mammal}
species      &   sample              & \makecell[r]{mammalian \\ mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\ 
\multirow{28}{*}{raccoon dog} & HJ200050-20200112-1 &                                       94\% &                        0 &                     1e+08 \\
                  & HJ200023-20200112-1 &                                       93\% &                        0 &                   6.9e+07 \\
                  & HJ200048-20200112-1 &                                       92\% &                        0 &                   1.2e+08 \\
                  & HJ200019-20200112-1 &                                       90\% &                        0 &                     7e+07 \\
                  & {\bf Q61} &                                       {\bf 81\%} &                        {\bf 1} &                   {\bf 2.1e+08} \\
                  & HJ200017-20200112-1 &                                       81\% &                        0 &                   1.1e+08 \\

>>>>

To determine the relationship between the 696 FASTQ files released by the Chinese CDC on the NGDC on March-29-2023 and the 227 files downloaded earlier that month by \citet{crits2023genetic}, I computed the SHA-512 hashes for all the NGDC FASTQ files and compared them to the hashes reported in Table S1 of \citet{crits2023genetic}.
This analysis confirmed that the dataset released on the NGDC contained unmodified versions of all 227 FASTQ files downloaded by \citet{crits2023genetic}, plus an additional 469 FASTQ files (Tables S1 and S2).

\subsection{Analysis of mitochondrial sequences validates mammalian metagenomics by Crits-Christoph et al, and also finds abundant non-mammalian chordates}
To determine the metagenomic content of the samples, I performed an analysis that conceptually parallels that described by \citet{crits2023genetic}.
Throughout this paper, I report results only for sequencing described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the small number of samples that involve amplicon-based sequencing of SARS-CoV-2 or RNA sequencing for viral whole-genome assembly (Table S2).

Briefly, the deep sequencing data were pre-processed to retain only high-quality reads.
These reads were aligned to the SARS-CoV-2 genome and a representative set of chordate mitochondrial genomes.
%The set of mitochondrial genomes was assembled by downloading all NCBI RefSeq genomes, filtering for those from chordates, and removing genomes that were highly similar to other retained genomes (within a Mash distance of $<$0.07).
This set included all the mammalian mitochondrial genomes from all species for which alignments were reported in \citet{crits2023genetic}.
The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table S3.
The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables S4 and S5, respectively.

I compared the mitochondrial compositions of the samples determined by my analysis to those reported by \citet{crits2023genetic}.
Note that while \citet{crits2023genetic} describe aligning the sequencing data to all metazoa mitochondria, they only report alignment counts for mammals.
The aligned read counts for mammals from my analysis are highly correlated with those reported by \citet{crits2023genetic} for all sequencing runs with a reasonable number of total counts (Figure S1).
My analysis and that of \citet{crits2023genetic} are independently implemented, and use different alignment programs, reference genome sets, and quality filtering criteria---so the fact that the two analyses give highly correlated results is reassuring.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{fig1.pdf}
\caption{
Metagenomic composition of sample Q61.
{\bf (A)}
Composition as determined by aligning reads to mitochondrial genomes.
From left to right: composition determined in current study across all chordates, composition determined in current study across all mammals, composition reported in \citet{crits2023genetic} for just mammals shown in the first figure of their report, and composition reported \citet{crits2023genetic} across all mammals in the third supplementary table of their report.
{\bf (B)}
Composition determined in the current study by aligning assembled contigs to the four indicated genomes; this composition is similar to that reported in the third figure of \citet{crits2023genetic}.
See \url{https://jbloom.github.io/Huanan_market_samples/mito_composition.html} for an interactive version of (A) that allows similar pie charts to be viewed for any sample.
See \url{https://jbloom.github.io/Huanan_market_samples/genomic_contig_composition.html} for an interactive version of (B).
\label{fig:composition}
}
\end{figure*}

\subsection{Most environmental samples contain little or no SARS-CoV-2 reads}
To quantify the SARS-CoV-2 content of the samples, I plotted the percentage of all high-quality reads that aligned to SARS-CoV-2 (Figure~\ref{fig:sars2_aligned}, Table S6, Table S7).
Over three-quarters of the samples had no reads aligning to SARS-CoV-2, and most of the rest only had a small number of viral reads.

@@

\begin{figure*}
\centerline{\includegraphics[width=0.85\linewidth]{fig2.pdf}}
\caption{
Percentage of high-quality reads that align to SARS-CoV-2 for each sample.
Points are colored according to the date that the sample was collected.
Note the x-axis uses a symlog scale.
See \url{https://jbloom.github.io/Huanan_market_samples/sars2_aligned_vertical.html} for an interactive version of this plot where you can mouseover points for details including the mitochondrial composition of each sample, and select only samples from specific dates or from locations.
\label{fig:sars2_aligned}
}
\end{figure*}

@@

See Table S8 for a similar table that shows mammalian rather than chordate composition.
This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table S9 (for raccoon dog) and Table S5 (for all species).

@@

 Overall, there were SARS-CoV-2 reads in just 3 of the 28 samples with at least 20\% of their chordate mitochondrial composition from the non-human susceptible species thought to have been sold live in the market (Table~\ref{tab:susceptible}).
There were 14 samples with at least 20\% chordate mitochondrial composition from raccoon dogs, but only sample Q61 contained any SARS-CoV-2 reads (1 of $\sim2.1 \times 10^8$ high-quality reads).
None of the six samples with at least 20\% chordate mitochondrial composition from bamboo rats contained any SARS-CoV-2 reads.
There was one sample each with at least 20\% chordate mitochondrial composition from Amur hedgehog and Malayan porcupine that contained SARS-CoV-2 reads (Table~\ref{tab:susceptible}).
If mitochondrial composition is instead analyzed only among mammals rather than chordates, Q61 is still the only sample with at least 20\% raccoon dog composition that contains any SARS-CoV-2 reads (Table S8).
There is one sample with at least 20\% of its mammalian (but not chordate) mitochondrial composition from bamboo rat that contains a small number of SARS-CoV-2 reads (Table S8).

Note that the 20\% cutoff applied in Tables~\ref{tab:susceptible} and S8 is not integral to the analysis, but is just a way to subset on samples containing the most material from the species of interest to make the tables small enough to easily visualize.
See Tables S9 and S5 for much larger tables that show comparable data for all samples with respect to the raccoon dog or overall mitochondrial composition, respectively.
For instance, Table S9 shows that there are a few samples with lower raccoon dog mitochondrial composition that also contain some SARS-CoV-2 reads.
However, ultimately these very large tables are difficult to visualize, and so the relationship between SARS-CoV-2 and mitochondrial content for each species for all samples is probably more facilely visualized using the interactive versions of the scatter plots described in the next subsection.

@@

Furthermore, the correlations are highly contingent on the sample set and details of how the statistics are calculated.
For instance, if we exclude the first sampling timepoint and only look at samples calculated on January-12-2020 or later, then the most correlated species is the Oriental rat snake, with the correlation for humans becoming even weaker and that for raccoon dogs remaining negative (Figure S2).
The correlations also change if only consider samples containing a non-zero number of SARS-CoV-2 reads, or calculate correlations on a linear rather than log scale (Figure S3).
The interactive versions of the correlation plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} and \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} enable the reader to explore the correlations for different sample subsets and methods of calculating the statistics.
%Exploration of these interactive plots shows that it takes a great deal of cherry picking of parameter combinations to put any plausibly infectable hosts at the top of the list of species whose mitochondrial material is most correlated with the abundance of SARS-CoV-2.

\begin{figure*}
\includegraphics[width=\linewidth]{fig3.png}
\caption{
Correlation between percent of all reads mapping to SARS-CoV-2 and the mitochondrial genome of each of the indicated species.
Each point represents a different environmental sample, and the orange text shows the Pearson correlation.
The scales are log10, and values of zero (which cannot be plotted on a log scale) are shown as half the minimum non-zero value observed across all samples.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot that enables mouseover of points for sample details, selection only of samples collected on specific dates or containing at least one SARS-CoV-2 reads, adjustment of scales from log to linear, and adjustment of mitochondrial percent to be of reads mapping to any mitochondria rather than of all reads.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for similar plots for individual species.
The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
\label{fig:per_species_corr_faceted}
}
\end{figure*}

@@

\begin{figure}
\includegraphics[width=\linewidth]{fig4.pdf}
\caption{
Correlations between SARS-CoV-2 content and mitochondrial content for all species.
Top row show just samples containing at least one SARS-CoV-2 read and bottom row shows samples regardless of SARS-CoV-2 content; left shows samples from all sampling dates and right shows just samples from the January-12-2020 date when most of the wildlife sampling occurred.
This plot is designed to mimic the fourth figure of \citet{liu2022surveillance}.
Some key species are labeled; see the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
See Figure S4 for a version of this plot that uses correlations based on a Theil-Sen estimator that is more robust to outliers.
\label{fig:overall_corr}
}
\end{figure}

To remedy this omission, I used the chordate mitochondrial compositions calculated in the current study to generate plots analogous to that in the original Chinese CDC preprint.
When the correlations are taken across all sampling dates only for samples containing at least one SARS-CoV-2 read (as done in the original Chinese CDC preprint), then the genetic material of several non-human species (largemouth bass, catfish, cow, sheep, pig) is more correlated with SARS-CoV-2 reads than is human material (Figure~\ref{fig:overall_corr}, upper left).
If the plot is expanded to all samples (regardless of whether or not they contain SARS-CoV-2 reads), the trends are broadly similar, with fish remaining the most correlated species (Figure~\ref{fig:overall_corr}, lower left).
It is also informative to take the correlations only across samples collected on January-12-2020, which was the date of most intensive sampling of the wildlife stalls (this is when Q61 was collected).
 For January-12-2020 samples that contain SARS-CoV-2 reads, humans are among the species whose genetic material is most correlated with SARS-CoV-2 reads, but goat and spotted dove have roughly equivalent correlations (Figure~\ref{fig:overall_corr}, upper right).
 If we include January-12-2020 samples regardless of SARS-CoV-2 content, then snakehead fish and Malayan porcupine are among the most correlated species.
Raccoon dog and bamboo rat genetic material are not positively correlated with SARS-CoV-2 reads in any of the sample sets.
See the interactive version of Figure~\ref{fig:overall_corr} at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to explore other sample subsets.
The trends are broadly similar if the correlations are instead calculated based on a Theil-Sen estimator~\citep{theilsen} that is more robust to outliers (Figure S4).

@@

Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2 (Table S10 gives sequencing and RT-qPCR data for all samples that were either sequenced or called positive by the Chinese CDC).
However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
This inconsistency is illustrated in Table~\ref{tab:consistency} for four example samples: F100 is clearly positive (it both tested positive by RT-qPCR and contained thousands of reads mapping to SARS-CoV-2), but there is no statistical rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative.

@@

For the current study, all mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
The Amur hedgehog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
All the mitochondrial genomes were then filtered to remove highly similar ones.
To do this, \texttt{Mash}~\citep{ondov2016mash} was used to compute the mash distances between all pairs of mitochondrial genomes.
To ensure the most relevant mitochondrial genomes were retained, I first specified for manual retention the genomes for all the species listed in the third supplementary file of \citet{crits2023genetic}, as well as genomes for some additional relevant species (these species are listed under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}).
After retaining all of these manually specified mitochondrial genomes, the pipeline then greedily iterated through all other genomes choosing for retention each genome that was not within a mash distance of 0.07 of another already retained genome.
The full set of retained mitochondrial genomes is listed in Table S3.

@@

I then used \texttt{CoverM} (\url{https://github.com/wwood/CoverM}) to compute the number of aligned reads for each run, requiring alignment lengths of at least 40 nucleotides with at least 95\% identity, and excluding the 100 nucleotides at the contig ends.
The resulting statistics on alignment counts to SARS-CoV-2 and the mitochondrial genomes for each sequencing run are in Tables S4 and S6.
To get per-sample alignment counts, I aggregated counts for all runs for each sample to get the statistics in Tables S5 and S7.
For the mitochondrial genome compositions and all other analyses in this paper, I only included only the samples from metagenomic sequencing, which are annotated with the description ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' in the metadata provided by \citet{liu2023surveillance}.
Species were only retained in the annotated reference set if they had at least 20\% of all aligned mitochondrial reads and at least 4000 covered based for at least one run, or if they are one of the species specified under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}.
The species specified under that key include all susceptible species studied in \citet{crits2023genetic}, so this criteria does not lead too dropping of any of the species in their study.

@@
